\MakeUppercase{The Structure of $k$-Planar Graphs}

### 2019-07-09 k-planar.tex
History of implicit representations

Here $\boxtimes$ is the strong product, and treewidth is an invariant that measures how `tree-like' a given graph is; see \cref{Tools} for formal definitions and see \cref{ProductExample} for an example. Loosely speaking, \cref{PlanarBasic} says that every planar graph is contained in the product of a tree-like graph and a path. This enables combinatorial results for graphs of bounded 
\begin{wrapfigure}{r}{0.5\textwidth} 
\centering
\includegraphics{ProductExample}
\caption{Example of a strong product.}
\label{ProductExample}
\end{wrapfigure}
treewidth to be generalised for planar graphs (with different constants). \cref{PlanarBasic} has been the key tool in solving several well-known open problems. In particular, \citet{dujmovic.joret.ea:planar} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92} from 1992), \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02} from 2002), \citet{bonamy.gavoille.ea:shorter} use it to find shorter implicit representations of planar graphs.

@@

\begin{figure}
  \begin{center}
    \includegraphics{figs/tripoddo}
  \end{center}
  \caption{The sets $Y_x$, $F_x$, and $N_x$ associated with $x\in V(T)$
  and the ancestors $a_1,\ldots,a_{t'}$ of $X$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.}
  \label{fig:generalized-tripod}
\end{figure}

@@

\begin{proof}
  We will count the number of vertices in $S_x\cap V_i$ by upper-bounding the number of vertices contributed to $S_x\cap V_i$ by each vertex $w\in Y_x$. Refer to \cref{contribute}.
  If $w\in Y_x\cap V_i$ and no path in $\mathcal{P}$ includes $w$ as an internal vertex then $w$ contributes only one vertex, itself, to $S_x\cap V_i$.
  \begin{figure}
    \begin{center}
      \includegraphics{figs/contribute}
    \end{center}
    \caption{A path $P$ containing an internal vertex $w\in Y_x\cap V_{i-j}$.}
    \label{contribute}
  \end{figure}

@@

The following lemma is due to \citet{micek:personal}. We include the proof  for completeness. 
% \note{DW}{can we replace by a result of \citet{micek:personal}?}
% \note{PM}{I didn't see a statement of this exact result in the last draft Piotr sent.  Though their Theorem~7 is this proof for the special case of planar graphs.}

\begin{lem}[\citep{micek:personal}]
\label{p-centered}
Every graph $G$ that has an $H$-partition of layered width at most $\ell$ in which $H$ has treewidth at most $t$ satisfies $\chi_p(G)\le \ell (p+1) (t+1)\binom{p+t}{t}$. 
\end{lem}

\begin{proof}
  Let $(\mathcal{L}=\langle V_0,V_1,\ldots\rangle, \mathcal{P}=(B_x:x\in V(H))$ be an $H$-partition of $G$ having layered width at most $\ell$ where $H$ as treewidth at most $t$.
  We use a product colouring $\phi:V(G)\to \{1,\ldots,\ell\}\times\{0,\ldots,p\}\times\{1,\ldots,
  (t+1)\binom{p+t}{t}\}$.  For each integer $i\ge 0$ and each $x\in V(H)$, assign the colour $\phi(v):=(\alpha(v),\beta(v),\gamma(v))$ to each vertex $v\in V_i\cap B_x$ such that:
  \begin{compactenum}
    \item $\alpha(v)$ is unique among $\{\phi(w): w\in V_i\cap B_x\}$, which is possible 
    since $|V_i\cap B_x|\le \ell$,
    \item $\beta(v)= i\bmod (p+1)$, and     
    \item $\gamma(v)=\gamma(x)$ where $\gamma:V(H)\to\{1,\ldots,(t+1)\binom{p+t}{t}\}$ is a $p$-centered colouring of $H$ from \cref{p-centered-treewidth}.
  \end{compactenum}
To show this is a $p$-centered colouring, consider some connected subgraph $X\subseteq G$.

First suppose that there exists $v,w\in V(X)$ with $v\in V_i$ and $w\in V_j$ with $j-i\ge p$. Since $G[X]$ is connected, $G[X]$ contains a path from $v$ to $w$.  By the definition of layering, this path contains at least one vertex from $V_{i'}$ for each $i'\in\{i,i+1,\ldots,j\}$. Therefore, $|\{\beta(v'):v'\in X\}|\ge j-i+1 > p$, so $X$ receives more than $p$ distinct colours.

Otherwise, $X\subseteq V_{i},\ldots,V_{i+s}$ for some $s<p$.  Let $H':=H[\{x\in V(H):B_x\cap X\neq\emptyset]$.  If $|\{\gamma(x):x\in V(H')\}| > p$ then $|\{\gamma(v):v\in X\}|> p$ so $|\{\phi(v):v\in X\}|> p$ and we are done.  Otherwise, since $\gamma$ is a $p$-centered colouring of $H$, there must exist some $x\in V(H')$ such that $\gamma(x)\neq\gamma(y)$ for every $y\in V(H')\setminus\{x\}$.
    For any $v,w\in B_x$ with $v\neq w$, either $v,w\in V_{i'}$ for some $i'\in\{i,i+1,\ldots,i+s\}$ in which case $\alpha(v)\neq\alpha(w)$; or $v\in V_{i'}$ and $w\in V_{i''}$ with $|i'-i''|< p$, in which case $\beta(v)\neq\beta(w)$. Therefore every vertex $v\in B_x$ receives a colour $\phi(v)$ distinct from every colour in $\{\phi(z):z\in X\setminus\{x\}\}$. Therefore, every vertex in $B_x$ receives colour distinct from every other vertex in $X$.\end{proof}

\cref{p-centered,k-planar,1-planar,g-k-planar} immediately imply the following results, for every $p\ge 2$:

>>>>

Here $\boxtimes$ is the strong product, and treewidth is an invariant that measures how `tree-like' a given graph is; see \cref{Tools} for formal definitions and see \cref{ProductExample} for an example. Loosely speaking, \cref{PlanarBasic} says that every planar graph is contained in the product of a tree-like graph and a path. This enables combinatorial results for graphs of bounded 
treewidth to be generalised for planar graphs (with different constants). \cref{PlanarBasic} has been the key tool in solving several well-known open problems. In particular, \citet{dujmovic.joret.ea:planar} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92} from 1992), \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02} from 2002), and \citet{bonamy.gavoille.ea:shorter} use it to find shorter implicit representations of planar graphs (making progress on a sequence of results going back to 1988 \citep{kannan.naor.ea:implicit-stoc,kannan.naor.ea:implicit}).

% \begin{wrapfigure}{r}{0.5\textwidth} 
\begin{figure}[hbtp]
\begin{center}
\includegraphics{ProductExample}
\end{center}
\caption{Example of a strong product.}
\label{ProductExample}
\end{figure}
% \end{wrapfigure}

@@

\begin{figure}[htbp]
  \begin{center}
    \includegraphics{figs/tripoddo}
  \end{center}
  \caption{The sets $Y_x$, $F_x$, and $N_x$ associated with $x\in V(T)$
  and the ancestors $a_1,\ldots,a_{t'}$ of $X$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.}
  \label{fig:generalized-tripod}
\end{figure}

@@

\begin{proof}
  We will count the number of vertices in $S_x\cap V_i$ by upper-bounding the number of vertices contributed to $S_x\cap V_i$ by each vertex $w\in Y_x$. Refer to \cref{contribute}.
  If $w\in Y_x\cap V_i$ and no path in $\mathcal{P}$ includes $w$ as an internal vertex then $w$ contributes only one vertex, itself, to $S_x\cap V_i$.

  \begin{figure}[htbp]
    \begin{center}
      \includegraphics{figs/contribute}
    \end{center}
    \caption{A path $P$ containing an internal vertex $w\in Y_x\cap V_{i-j}$.}
    \label{contribute}
  \end{figure}

@@

The following lemma is due to \citet{micek:personal}. 
% We include the proof  for completeness. 
% \note{DW}{can we replace by a result of \citet{micek:personal}?}
% \note{PM}{I didn't see a statement of this exact result in the last draft Piotr sent.  Though their Theorem~7 is this proof for the special case of planar graphs.}

\begin{lem}[\citep{micek:personal}]
\label{p-centered}
Every graph $G$ that has an $H$-partition of layered width at most $\ell$ has $\chi_p(G)\le \ell (p+1)\chi_p(H)$. 
\end{lem}

% \begin{proof}
%   Let $(\mathcal{L}=\langle V_0,V_1,\ldots\rangle, \mathcal{P}=(B_x:x\in V(H))$ be an $H$-partition of $G$ having layered width at most $\ell$.
%   Use a product colouring $\phi:V(G)\to \{1,\ldots,\ell\}\times\{0,\ldots,p\}\times\{1,\ldots,
%   (t+1)\binom{p+t}{t}\}$.  For each integer $i\ge 0$ and each $x\in V(H)$, assign the colour $\phi(v):=(\alpha(v),\beta(v),\gamma(v))$ to each vertex $v\in V_i\cap B_x$ such that:
%   \begin{compactenum}
%     \item $\alpha(v)$ is unique among $\{\phi(w): w\in V_i\cap B_x\}$, which is possible 
%     since $|V_i\cap B_x|\le \ell$,
%     \item $\beta(v)= i\bmod (p+1)$, and     
%     \item $\gamma(v)=\gamma(x)$ where $\gamma:V(H)\to\{1,\ldots,\chi_p(H)\}$ is a $p$-centered colouring of $H$.
%   \end{compactenum}
% To show this is a $p$-centered colouring, consider some connected subgraph $X\subseteq G$.
% 
% First suppose that there exists $v,w\in V(X)$ with $v\in V_i$ and $w\in V_j$ with $j-i\ge p$. Since $G[X]$ is connected, $G[X]$ contains a path from $v$ to $w$.  By the definition of layering, this path contains at least one vertex from $V_{i'}$ for each $i'\in\{i,i+1,\ldots,j\}$. Therefore, $|\{\beta(v'):v'\in X\}|\ge j-i+1 > p$, so $X$ receives more than $p$ distinct colours.
% 
% Otherwise, $X\subseteq V_{i},\ldots,V_{i+s}$ for some $s<p$.  Let $H':=H[\{x\in V(H):B_x\cap X\neq\emptyset]$.  If $|\{\gamma(x):x\in V(H')\}| > p$ then $|\{\gamma(v):v\in X\}|> p$ so $|\{\phi(v):v\in X\}|> p$ and we are done.  Otherwise, since $\gamma$ is a $p$-centered colouring of $H$, there must exist some $x\in V(H')$ such that $\gamma(x)\neq\gamma(y)$ for every $y\in V(H')\setminus\{x\}$.
% For any $v,w\in B_x$ with $v\neq w$, either $v,w\in V_{i'}$ for some $i'\in\{i,i+1,\ldots,i+s\}$ in which case $\alpha(v)\neq\alpha(w)$; or $v\in V_{i'}$ and $w\in V_{i''}$ with $|i'-i''|< p$, in which case $\beta(v)\neq\beta(w)$. Therefore every vertex $v\in B_x$ receives a colour $\phi(v)$ distinct from every colour in $\{\phi(z):z\in X\setminus\{x\}\}$. Therefore, every vertex in $B_x$ receives a colour distinct from every other vertex in $X$.
% \end{proof}

\cref{p-centered,p-centered-treewidth,k-planar,1-planar,g-k-planar} immediately imply the following results, for every $p\ge 2$:
